Instability and inequality in the COVID-19 lockdown crime drop

### 2021-03-09 draft_paper.Rmd
heat maps and writing

Recent months have seen a wave of interest in describing and explaining changes in crime in response to restrictions on mobility brought about to limit the spread of COVID-19. Studies include case study sites in the United States [@campedelli2020exploring; @gerell2020minor; @mohler2020impact; @piquero2020staying; @ashby2020initial], Australia [@payne2020covid; @payne2020property], the United Kingdom [@langton2021cs; @halford2020crime; @buil2020cybercrime] Canada [@hodgkinson2020show] and Mexico [@estevez2020crime]. Generally speaking, findings are consistent with opportunity perspectives and the mobility theory on the pandemic [@langton2021cs]. That is, legally-enforced restrictions on people's mobility and social interactions limited opportunities for criminality which in turn resulted in a macro-level decline in crime. As restrictions were relaxed, these opportunities reemerged, mobility patterns converged back to normality, and crime began to 'bounce back' -- often to levels that would have been expected without the global pandemic and ensuing lockdowns.

That said, little is known about the localized drivers of these macro-level trends. An exploration of the extent to which the macro lockdown crime drop has been experienced equitably across space is valuable for two principal reasons. First, we would expect only a small number of fine-grained geographic units to have driven the lockdown crime drop. In pre-pandemic era, longitudinal studies have consistently demonstrated that, in a crime drop scenario, most local areas remained remarkably stable, with a disproportionatly large volume of change attributable to just handful of meso or micro-level units (e.g. @weisburd2004trajectories; @akmed2021mon; @curman2015crime; @bannister2018local). Findings have contributed to the evidence-base around routine activities theory, reflecing the considerable variation in criminal opportunities across (even spatially proximate) places [@eck2000finding; @weisburd2012criminology].

Secondly,

from an opportunity and mobility perspective,

preliminary evidence from the United States suggests that macro lockdown crime trends might mask underlying variation. Using police-recorded crime data in San Francisco and Oakland, @shayegh2020staying provided visual evidence indicating geographic variation in that the fall in crime rates ac

Secondly, existing studies suggest small number of drivers.

Thirdly, equity argument.

@shayegh2020staying; 
@felson2020routine

- Findings hold important practical and theoretical value for crime science research, and yet, less is known about the localized drivers of macro-level trends. 
- Questions around the drivers of macro-level crime drops have been of interest for some time, following the widespread fall in crime since the 1990s [@farrell2014crime; @weisburd2012criminology; @weisburd2004trajectories; @bannister2018local].
- At the city level, numerous studies have deployed longitudinal clustering methods to unmask localized variation in the crime drop. Findings have consistently found that macro-level declines in crime over many years (decades) tend to be driven by a disproportionately small number of places, with most areas remaining fairly stable.
- Here, we are presented with a unique scenario. Rather than a gradual decline in police-recorded crime over many decades, we have witnessed a rapid decline over a matter of weeks and months. And yet, comparable questions remain.
- To what extent have local areas experienced this decline in unison?
- Can we attribute most of the macro-level count trends to a handful of meso-level areas? 
- Have the dividends of the lockdown crime drop been shared equitably across space [@akmed2021mon]?
- Here, we use 7-months of police-recorded crime in England and Wales, aggregated to meso-level local units, to disentangle the nationwide lockdown crime drop.
- We describe areas in terms of their opportunity structure, spatial patterning and crime type profile.

@@

We deploy k-means using the `kml` package [@kmlpack] in R on notifiable offences (excluding drugs) counts for LSOA in the study region (N = 33,075) between February and August 2020. To achieve a parsimonious cluster solution while minimizing the risk of missing underlying variation, we proposed potential solutions between two and eight clusters respectively, choosing the final solution based on the Calinski-Criterion [@calinski1974dendrite]. For each potential solution, twenty redraws with different starting conditions were run to ensure that solutions were stable (see Appendix).

We visualize the final cluster solution in a manner which conveys the underlying distribution of observations comprising each cluster at each time point, rather than reporting a summary statistic (e.g. the mean trajectory) in isolation, as has been in the case in existing research. For each cluster, we also overlay the equivalent trajectories for 2018 and 2019 as a reference point for comparison to a 'typical' year. In doing so, we aim to not only identify localized (in)stability in the lockdown crime drop, but also assess the extent to which the trends observed have deviated from previous years.

@@

We deploy non-parametric k-means clustering on LSOA (meso) level geographic units (N = 33,075) comprising England and Wales to decompose the macro-level 'decline and resurgence' observed between February and August, 2020. Figure 3 visualizes the resulting 6-cluster solution determined by the optimal Calinski-Criterion statistic [@calinski1974dendrite]. Solid black lines represent the median count for each cluster at any given time point, with the black dotted line showing the mean. Violin plots have been added to convey variation around these points at each time point. These suggest that the mean and median point statistics summarize the underlying data reasonably well, and indicate that clusters are distinct from one another, with little overlap. Additional lines have been added to convey each clusters' mean and median trend in 2018 and 2019 respectively. These trends suggest that the clusters identified using the 2020 study period were distinct and meaningful even in previous years, and provide a relative baseline from which we can compare lockdown trends.

@@

That said, most 'action' appears to be occurring amongst a small subset of LSOA. Clusters E and F collectively comprise only 0.33% (N = 110) of LSOA in England and Wales. Yet, their crime counts are much higher, and the decline between March and April is considerable. We might therefore describe these clusters as 'high crime, major drop': LSOA with plentiful opportunities for crime in typical times, and in turn, LSOA which are most sensitive to the restriction in opportunities which followed after the imposition of lockdown. Interestingly, clusters E and F represent the only clusters which remained below pre-pandemic levels by August. In both cases, the average counts in February 2020 were _higher_ than those observed in previous years, but following the sharp decline and subsequent increase by August, these averages were _lower_ than 2018 and 2019.

@@

To further unpick the contribution of these clusters to the nationwide lockdown crime drop, Figure 4 plots the month-by-month change in counts. Alongside these raw counts, we also include the percentage of total absolute change (i.e. positive or negative) attributable to each cluster. We can use this visual to identify which clusters drove the initial decline and subsequent resurgence nationwide.

@@

The 'low crime and stable' clusters (A and B) actually _declined again_ between July and August, and the higher crime clusters continued to increase. The change is almost imperceptible when viewing average counts in these LSOA (see Figure 3) but the sheer size of these units, collectively comprising 89% of England and Wales, has a major impact on nationwide trends. This is suggestive of highly localized change in opportunities as lockdown rules were relaxed: change which is aggregated away at the macro-level. Stable, low crime areas mitigated against further nationwide increases which may, if they had continued, actually resulted in a higher crime count in August than observed in previous years.

@@

>>>>

Recent months have seen a wave of interest in describing and explaining changes in crime in response to restrictions on mobility brought about to limit the spread of COVID-19. Studies include case study sites in the United States [@campedelli2020exploring; @gerell2020minor; @mohler2020impact; @piquero2020staying; @ashby2020initial], Australia [@payne2020covid; @payne2020property], the United Kingdom [@langton2021cs; @halford2020crime; @buil2020cybercrime] Canada [@hodgkinson2020show] and Mexico [@estevez2020crime]. Generally speaking, findings are consistent with opportunity perspectives and the mobility theory of crime during the pandemic [@langton2021cs; @halford2020crime]. That is, legally-enforced restrictions on people's mobility and social interactions limited opportunities for criminality which in turn resulted in a macro-level decline in crime. As restrictions were relaxed, these opportunities reemerged, mobility patterns converged back to normality, and crime began to 'bounce back' -- often to levels that would have been expected without the global pandemic and ensuing lockdowns.

That said, little is known about the localized drivers of these macro-level trends. To what extent have local areas experienced the lockdown crime drop in unison? This exploration represents a unique test of opportunity perspectives on crime, which would stipulate that only a small number of local areas will have driven the lockdown crime drop. In the pre-pandemic era, longitudinal studies have consistently demonstrated that, in a macro-level crime drop scenario, most meso or micro-level areas remained remarkably stable, with a disproportionately large volume of change attributable to just handful of units [e.g. @weisburd2004trajectories; @akmed2021mon; @curman2015crime; @bannister2018local]. This is credited to highly localized opportunity structures, which can differ considerably within the same city, and even within the same neighborhood [@eck2000finding; @weisburd2012criminology]. Here, we expect that areas typically characterized by plentiful opportunities for crime, which quickly become unavailable following the imposition of lockdown measures, will drive the lockdown crime drop.

Preliminary evidence from the United States certainly suggests that macro lockdown crime trends might mask underlying variation. Using police-recorded crime data in San Francisco and Oakland, @shayegh2020staying provided visual descriptive evidence which indicated some degree of geographic variation in pre and post-lockdown crime. A study of Detroit census block groups found that, amidst a fall in burglaries following a stay at home order, there was a shift in concentrations away from residential areas towards mixed and non-residential parts of the city [@felson2020routine].

- We describe areas in terms of their opportunity structure, spatial patterning and crime type profile.

@@

We deploy k-means using the `kml` package [@kmlpack] in R on notifiable offences (excluding drugs) counts for LSOA in the study region (N = 33,075) between February and August 2020. To achieve a parsimonious cluster solution while minimizing the risk of missing underlying variation, we proposed potential solutions between two and eight clusters respectively, choosing the final solution based on the Calinski-Criterion [@calinski1974dendrite]. For each potential solution, twenty redraws with different starting conditions were run to ensure that solutions were stable. Potential cluster solution options were also examined using Principal Component Analysis to establish their suitability.

We visualize the final cluster solution in a manner which conveys the underlying distribution of observations comprising each cluster at each time point, rather than reporting a summary statistic (e.g. the mean trajectory) in isolation. For each cluster, we also overlay the equivalent trajectories for 2018 and 2019 as a reference point for comparison to a 'typical' year. In doing so, we aim to not only identify localized (in)stability in the lockdown crime drop, but also assess the extent to which the trends observed have deviated from previous years.

@@

We deploy non-parametric k-means clustering on LSOA (meso) level geographic units (N = 33,075) comprising England and Wales to decompose the macro-level 'decline and resurgence' observed between February and August, 2020. Based on an assessment of the Calinski-Criterion statistic [@calinski1974dendrite] and Principal Component Analysis (see Appendix), we determined the optimal solution to be 6 clusters (see Figure 3). Solid black lines represent the median count for each cluster at any given time point, with the black dotted line showing the mean. Violin plots have been added to convey variation around these points at each time point. These suggest that the mean and median point statistics summarize the underlying data reasonably well, and indicate that clusters are distinct from one another, with little overlap. Additional lines have been added to convey each clusters' mean and median trend in 2018 and 2019 respectively. These trends suggest that the clusters identified using the 2020 study period were distinct and meaningful even in previous years, and provide a relative baseline from which we can compare lockdown trends.

@@

That said, most 'action' appears to be occurring amongst a small subset of LSOA. Clusters E and F collectively comprise only 0.33% (N = 110). Yet, their crime counts are much higher, and the decline between March and April is considerable. We might therefore describe these clusters as 'high crime, major drop': LSOA with plentiful opportunities for crime in typical times, and in turn, LSOA which are most sensitive to the restriction in opportunities which followed after the imposition of lockdown. To further decompose these clusters, we now disentangle the contribution of each cluster.

@@

To further unpick the contribution of these clusters to the nationwide lockdown crime drop, Figure 4 plots the monthly change in counts and percentage of total absolute change (i.e. positive or negative) attributable to each cluster. We can use this visual to identify which clusters drove the initial decline and subsequent nationwide resurgence in crime.

@@

The 'low crime and stable' clusters (A and B) actually _declined again_ between July and August, and the higher crime clusters continued to increase. The change is almost imperceptible when viewing average counts in these LSOA (see Figure 3) but the sheer size of these units, collectively comprising 89% of England and Wales, has a major impact on nationwide trends. This is suggestive of highly localized change in opportunities as lockdown rules were relaxed: change which is aggregated away at the macro-level. Stable, low crime areas mitigated against further nationwide increases which may, if they had continued, resulted in a higher crime count in August than observed in previous years.

@@

These findings can be supplemented with a comparison to the previous year. Figure 5 compares the clusters identified in 2020 with the equivalent months in 2019. Here, it is clear that the 'high crime major drop' clusters (E and F) contributed the most to the lockdown crime drop relative to their size. By way of example, around 27,000 fewer crimes occurred in cluster A during April 2020 compared to April 2019 (Figure 5a). This translates to a 24% drop (Figure 5b), and an average fall per LSOA of 1.39 counts (Figure 5c). For cluster F, approximately 5,000 fewer crimes occurred (Figure 5a), which translates to an 87% drop (Figure 5b), and an average fall per LSOA of 543 counts (Figure 5c). In other words: whether comparing month-to-month change in 2020, or comparing equivalent months in 2019, the lockdown crime drop was driven by a small number of local areas, with most meso-level units remaining fairly stable, even amidst dramatic macro-level change.

```{r heatmaps, fig.pos = 'H', message=F, comment=F, warning=F, fig.cap = "Comparing clusters between equivalent months in 2020 and 2019 according to (a) absolute difference in counts, (b) percentage difference in counts, (c) average difference in counts.", fig.align='center'}
knitr::include_graphics("visuals/heatmaps_gg.png")
```
